# Recommendations
The Neexdzii Kwah watershed  has been extensively impacted not only by physical infrastructure, such as the linear corridors of CN Rail, Highway 16, resource extraction roads, and utility lines, but also by the legacy of colonization. Recovery efforts must go beyond technical interventions and include a commitment to learning from the vast body of knowledge held by the Wetsuwet'en people. This includes understanding cultural practices, place names, family connections, house groups, poles, and feasts. There is extensive literature documenting these histories, including records from the Delgamuukw trial, which provide a foundation for understanding the cultural significance of this landscape.

<br>

Developing comprehensive restoration planning for the Neexdzii Kwah watershed will help ensure that the restoration activities address the underlying drivers of the degradation necessitating restoration. Through partnerships, education, and resource-sharing, stakeholders can work together to align ecological goals with economic interests, ensuring long-term benefits for the watershed. The following recommendations are provided to try to ensure that the restoration activities are most effective by addressing the underlying drivers of the degradation. At the time of writing this can be considered a partial list as the project is still in development and the recommendations are subject to change as the project progresses:

```{r, echo=FALSE}
knitr::include_url("https://www.newgraphenvironment.com/restoration_wedzin_kwa_2024_recomendations/", height = "800px")
```


```{r}
# 1.  Amalgamate effectiveness monitoring information and present within online reporting to facilitate adaptive management through sharing and review. Evolve program to leverage past monitoring methodology and data include an updated monitoring schedule, monitoring objectives, and methods. Methodology using Watershed Restoration Procedures - Routine effectiveness evaluation procedures and riparian inventory techniques ([LMH25](https://www.for.gov.bc.ca/hfd/pubs/Docs/Lmh/Lmh25/Lmh25_ed2_(2010).pdf) - percent cover by structural stage, species composition, etc.) could be utilized using pre-built digital forms served to field teams from the collaborative GIS project.
# 
# 1.  Design long term water monitoring program to leverage past work and attempt to quantify upstream water quality impacts via tools such as [CABIN](https://www.canada.ca/en/environment-climate-change/services/canadian-aquatic-biomonitoring-network/resources.html) sampling program [@canada2008CanadianAquatic], 5 sample in 30 day water quality sampling during both high and low flow periods [@wlrs2024BritishColumbia; @moe2024BritishColumbia], continuation/expansion of [temperature monitoring](https://public.tableau.com/app/profile/skeena.knowledge.trust/viz/UBRWaterTemperatureMonitoringDashboardDraft/UBRWaterTemp_Dashboard) [@skeenaknowledgetrustUBRWater, @westcott2022UpperBulkleya], quantify amounts of water withdrawn by licensees in the upper Bulkley catchment during the April to September low flow period, etc).  Importantly, careful consideration of how to implement water quality program recommendations from  @price2014UpperBulkleya, @oliver2020Analysis2017 and @westcott2022UpperBulkleya during this process is also recommended.
# 
# 
# 2. Incentivize and help coordinate range tenure practice improvements (off-site watering, rotational field use, riparian exclusion, monitoring, range use plan updates that reduce cattle load, range stewardship planning that includes "red" zones for cattle exclusion, etc.) to reduce cattle impacts on riparian and water quality values.  Cattle waste loading to tributary and mainstem habitats is of paramount concern as water quality degradation due to high biological oxygen demand related to nutrient loading will undermine the ability of aquatic life to recover from cumulative impacts in the watershed.
# 
# 3.  Leverage and expand existing education programs for youth related to watershed health and cultural/ecosystem significance of fish/riparian/wildlife values to provide outreach for all school aged and post-secondary students in the greater Skeena region.
# 
# 4.  Work with organizations such as the Woodmere Nursery, Skeena based "Seed the North", knowledge holders and others to develop a local seed collection and propagation program to increase capacity for floodplain revegetation that can be scaled up to meet the challenges faced by historic removal of these native vegetation communities.
# 
# 10. In our recent field assessments, we observed that the protection of road and rail infrastructure through streambank armoring is not adequately incorporating best practices for vegetating riprap, soft armouring where possible and establishing/restoring effective riparian buffers. This oversight can significantly impact watershed health and biodiversity. To address this issue, we recommend collaborating with relevant stakeholders to advocate for the documentation, evolution, and dissemination of best practices related to streambank armoring.
# 
# 4.  Develop a schedule for on the ground monitoring via pre-planned instream surveys (walks) of mainstem and tributary habitats and riparian recovery/monitoring. Map these "trail" systems and document values within them. Increasing the presence of First Nation representatives, environmental practitioners and regulators within valuable streams within the Neexdzi Kwa watershed - regardless of land ownership - will help educate surveyors on the current state of these systems, provide information regarding recovery effort results and incentivize responsible land management by landowners/managers in the region.
# 
# 
# 6.  Select and refine parameters to prioritize future recovery program type (ex. water quality monitoring, range management program, education initiative, riparian vegetation recovery, expanded seed collection/propogation program, erosion protection, etc.) and allocate available dollars accordingly.  After financial and capacity resources are delineated by recovery program type - select parameters, and weight a ranking system for each to inform prioritization of specific sites within each recovery program. For physical actions such as riparian fencing for cattle exclusion, riparian/floodplain vegetation recovery and erosion protection include prioritization criteria that incorporates the amount of riparian/floodplain area available for recovery (ie. the wider the buffer available for meaningful native floodplain vegetation community establishment -  the higher the priority for investment).  Utilize established standards when possible such as the [Riparian Areas Protection Regulation](https://www2.gov.bc.ca/gov/content/environment/plants-animals-ecosystems/fish/aquatic-habitat-management/riparian-areas-regulation) and the [Riparian Management Area Guidebook](https://www2.gov.bc.ca/gov/content/industry/forestry/managing-our-forest-resources/silviculture/silvicultural-systems/silviculture-guidebooks/riparian-management-area-guidebook) [@ministryofforests1995RiparianManagement].
# 
# 7. Working with watershed users—including industrial players like railways and highways, as well as landowners and range tenure holders—is essential for promoting water retention through beaver activity. The Beaver Restoration Assessment Tool (BRAT) can guide efforts by identifying areas where beaver activity can optimize water retention and flow regulation. Beavers are natural ecosystem engineers; their dams slow water flow, enhance groundwater recharge, and reduce flood risks, while also improving water quality by trapping sediments and filtering pollutants. Providing incentives to landowners and tenure holders to support water retention on the landscape can help catalyze beaver influenced wetlands throughout the watershed addressing potential land use conflicts while aligning ecological goals with economic interests.
# 
# 5. Acquire, georeference, archive (within SKT) and analyze historic aerial imagery for the Neexdzii Kwah watershed and compare with recent orthoimagery and/or LiDAR data to quantify historic changes in stream morphology and floodplain extent resulting in loss of quantity and quality of water and fish habitat while highlighting areas of historic dredging, realignment, and floodplain disconnections due to infrastructure, to guide future restoration efforts. Script all analysis when possible to facilitate reproducibility and sharing of the analysis.
# 
# 8.  Continue to develop and document data management workflows to leverage existing established data storage systems to retrieve data from and load to and build capacity for all interested to do the same. Established data storage systems include the [BC Data Catalogue](https://catalogue.data.gov.bc.ca/), the [Skeena Salmon Data Centre](https://data.skeenasalmon.info/), [NuSEDS-New Salmon Escapement Database System](https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6) and [bcfishpass](https://github.com/smnorris/bcfishpass/tree/main/data)
# 
# 
# 9.  Amalgamate drone acquired ortho-imagery gathered for sites in the Neexdzii Kwah watershed, store as Cloud Optimized Geotiffs on a cloud service provider and link to the imagery via the collaborative GIS project. Orthoimagery has been gathered for past monitoring of historic restoration sites as well as part of fish passage restoration planning efforts.
# 
# 
# 
# 
# 
# 12. Link historic and potential future sites to the freshwater atlas using the linear feature id of stream segments and conduct watershed drainage area analysis for individual stream segments (or portions of steram segments) so that this area can be tied to other GIS and user input parameters. Functions kept in [fwapg](https://github.com/smnorris/fwapg), [fwapgr](https://github.com/poissonconsulting/fwapgr), [fwatlasbc](https://github.com/poissonconsulting/fwatlasbc) and custom function developed for this project can be used. Results from this can be compared to other methods for determining project setbacks from the high water mark such as the [Riparian Areas Protection Regulation](https://www2.gov.bc.ca/gov/content/environment/plants-animals-ecosystems/fish/aquatic-habitat-management/riparian-areas-regulation) and the [Riparian Management Area Guidebook](https://www2.gov.bc.ca/gov/content/industry/forestry/managing-our-forest-resources/silviculture/silvicultural-systems/silviculture-guidebooks/riparian-management-area-guidebook) [@ministryofforests1995RiparianManagement].
# 
# 
# 14.  Spatially represent historic fisheries sites presented in Table \@ref(tab:tab-hist-sites) so the option is available that potential areas of restoration can be ranked according to their proximity to these sites.
# 
# 15.  Spatially delineate areas of high fisheries values based on known spawning and rearing habitat for fish species of interest. Where data gaps exist - develop study plans to understand where this high value habitat is and quantify its value for fish species of interest through standardized methodologies such as Fish and Fish Habitat Assessment Procedures [@johnstonFishHabitatAssessment1996]. For known spawning and rearing locations, load reviewed data into `bcfishpass` using the `user_habitat_classification.csv` file located [here](https://github.com/smnorris/bcfishpass/blob/main/data/user_habitat_classification.csv) so it can be pulled into the collaborative GIS project and queried to facilitate priority ranking of areas of high fisheries values.
```

