# Recommendations
The Neexdzii Kwah watershed  has been extensively impacted by land clearing/alterations/pollution related to agricultural 
activities and resource extraction as well as by the development of linear corridors of CN Rail, Highway 16, resource 
extraction roads, and utility infrastructure.  Furthermore the area
has also been impacted by the legacy of colonization. Recovery efforts must go beyond technical interventions and include a 
commitment to learning from the vast body of knowledge held by the Wetsuwet'en people. This includes understanding cultural 
practices, oral tradition (Kungax), Wet'suwet'en Law (Ink Nu'at'en), place names, family connections, house groups (Yihk), 
poles (Ts'an), the feast system (Bahlats), etc [@harris2011Yinkadinii]. There is extensive literature documenting these 
histories, including `r ngr::ngr_str_link_url(url_base = "https://open.library.ubc.ca/collections/delgamuukw", anchor_text = "records from the Delgamuukw & Gisday Wa land claims")`  and within the book "Niwhts’ide’ni Hibi’it’ën: The Ways Of Our Ancestors. Witsuwit’en History & Culture 
Throughout The Millennia" available through  `r ngr::ngr_str_link_url(url_base = "https://www.sd54.bc.ca/contact-us/", anchor_text = "School District #54")`
, which can begin to provide a foundation for understanding the cultural significance of this landscape [@universityofbritishcolumbialibraryDelgamuukwTrial; @morin2016NiwhtsideniHibiiten].

<br>

This report should be considered a living document, with ongoing revisions and updates tracked through version control.
We are humbled by the depth of knowledge that exists for this region, both through traditional knowledge and scientific study.
We acknowledge that this report is far from complete, and that the work of restoring and sustaining ecosystem values will
require many decades—if not centuries—of dedication. It is our hope that this effort can serve as a catalyst for positive
change and foster genuine, open collaboration. 

<br>

To support long-term watershed resilience in Neexdzii Kwah, we recommend a coordinated strategy that includes implementing a cost-effective, culturally grounded ecosystem health monitoring program; adapting range use practices to reduce cattle impacts on water quality and riparian habitats; expanding community education programs that center Wet’suwet’en knowledge and values; and restoring floodplain function through process based restoration including riparian revegetation, recovery of wetland areas and stream reconnection. These efforts draw on past recommendations and aim to align ecological science with Indigenous stewardship for a healthy sustainable watershed.

<br>

Developing comprehensive restoration planning for the Neexdzii Kwah watershed will help ensure that restoration activities address the underlying drivers of degradation. Through partnerships, education, and resource-sharing, stakeholders can work together to align ecological goals with economic interests, ensuring long-term benefits for the watershed. The following recommendations are provided to support effective restoration by addressing root causes. This is a partial list as the project is still in development — progress is tracked through [issues](https://github.com/NewGraphEnvironment/restoration_wedzin_kwa_2024/issues) and the [project changelog](https://github.com/NewGraphEnvironment/restoration_wedzin_kwa_2024/blob/main/NEWS.md).

## Research as Territorial Stewardship {-}

The research priorities identified in this report are acts of territorial stewardship — the work of understanding and caring for the land in ways that align with yintah. How these programmes are designed and implemented matters as much as what they measure. @harris2011Yinkadinii argues that health, ecosystems, social systems, and watershed management must be governed together. The priorities below reflect the biophysical face of that vision, but their success depends on implementation within a governance framework that centres Wet'suwet'en authority and community participation [@gottesfeld_etal2009CaseHistory]. Monitoring programmes that engage Wet'suwet'en community members in enumeration, habitat assessment, and restoration are not just better science — they are part of the journey toward recovery, for both giis and the people whose wellbeing is tied to them.

The United Nations Decade on Ecosystem Restoration (2021–2030) reinforces these principles through international standards of practice [@fao_etal2023Standardspractice]. Three commitments are central:

1. **Address root causes of degradation, not just symptoms.** Armouring an eroding streambank while the adjacent floodplain is still being actively cleared is symptom management. Root causes in the Neexdzii Kwah include ongoing floodplain vegetation removal, loss of wetlands and off-channel areas, stream channelization, and cattle impacts on riparian areas — alongside fishing pressure that exceeds what salmon populations can sustain [@winther_etal2024assessmentskeena; @lglltd_2025SkeenaRiver; @price_etal2026rebuildinggiis]. Addressing these means recovering floodplain function through revegetation, wetland restoration, and process-based approaches — while simultaneously reducing harvest to levels that allow rebuilt habitat to produce fish that survive to spawn.

2. **Benefit the people with the deepest connection to the ecosystem.** Restoration programmes must be designed and governed by the communities whose wellbeing is tied to the resource. For the Neexdzii Kwah, this means Wet'suwet'en-led programme design, monitoring, and decision-making — consistent with both the wiggus principle of interconnectedness between people and land and the United Nations Declaration on the Rights of Indigenous Peoples [@unitednationsgeneralassembly2007UnitedNations].

3. **Integrate governance across scales.** Effective restoration requires alignment between the people managing the land, the agencies regulating harvest, and the international frameworks governing ocean fisheries. Fragmented governance — where habitat restoration proceeds in one jurisdiction while overfishing continues in another — cannot succeed.

## Prioritization Framework {-}

### Governance Structure {-}

Effective restoration requires not just identifying what to do, but establishing who decides. We propose a four-tier governance model anchored in Wet'suwet'en governance:

1. **Watershed Stewardship Council (Decision Authority)** — Wet'suwet'en hereditary leadership and technical advisors. Sets restoration priorities, approves projects, and holds accountability. Grounded in Wet'suwet'en law (Ink Nu'at'en) and land stewardship responsibilities.

2. **Technical Working Group (Science & Planning)** — Biologists, hydrologists, GIS analysts, and restoration practitioners. Develops project designs, monitors outcomes, and adapts approaches. Translates community priorities into implementable plans.

3. **Implementation Partnerships (Delivery)** — Landowners, contractors, conservation groups, and government agencies. Execute approved projects within the governance framework, with project complexity matched to delivery capacity.

4. **Community Roundtable (Input & Accountability)** — Open forum for landowners, residents, and stakeholders. Provides local knowledge, flags concerns, and reviews progress to ensure transparency and broad engagement.


### Gates Before Scoring {-}

Before a site enters the scoring framework, it must pass three diagnostic gates. Sites that fail a gate are not rejected — they are routed to the appropriate queue (assessment, pressure reduction, or relationship building) rather than scored prematurely.

| Gate | Question | If No |
|------|----------|-------|
| **G1: Diagnostic Certainty** | Do we understand the problem well enough to prescribe a treatment? | Route to assessment queue |
| **G2: Active Degradation Clear** | Is the pressure that caused the problem still ongoing? | Address pressure first |
| **G3: Access & Willingness** | Is there a willing partner and legal access? | Park until social license exists |


### Scoring Principles {-}

Seven principles guide how projects are evaluated once they pass all gates:

1. **Root-cause alignment** — Does the project address why the system is degraded, or just treat symptoms? Bank armouring a reach while adjacent floodplain vegetation is being actively cleared and cattle continue to access riparian areas is symptom management. Understanding root causes requires investment in diagnostics — review of historic aerial imagery (available back to 1968), land cover change analysis, water quality monitoring, and fish population data — before prescribing treatments. Without meaningful riparian buffers and with adjacent areas seeing active vegetation removal, engineered interventions are unlikely to achieve lasting outcomes.

2. **Scale of benefit** — Is the project at a scale that matters for watershed function? Land cover change detection between 2017 and 2023 quantifies floodplain tree loss across sub-basins (Table \@ref(tab:tab-area-priority-cap)), and historic aerial imagery extending back to 1968 can now place recent trends in a multi-decadal context. These tools allow us to assess whether a proposed project addresses a reach-scale problem or a pattern that extends across the sub-basin — and to match the scale of investment to the scale of impact.

3. **Passive-first** — Can the system recover on its own if we stop the pressure? Removing a barrier to natural process (e.g., restoring riparian buffers, supporting beaver activity) before engineering a solution.

4. **Probability of success** — Is there a willing landowner, legal access, and realistic maintenance path? Great designs fail without social license.

5. **Wet'suwet'en stewardship** — Does the project align with Wet'suwet'en values and governance? Cultural significance is not a tiebreaker — it is a foundational criterion.

6. **Learning & accountability** — Does the project include monitoring that builds shared understanding of ecosystem health? Monitoring is not an add-on to restoration — it is itself an act of stewardship, restoring our collective understanding of what is happening on the land. Effectiveness monitoring, water quality sampling, fish population tracking, and land cover change analysis must be resourced as fundamental components of the work, not optional extras contingent on leftover budget. Without sustained investment in monitoring, we cannot learn from our interventions, adapt our approaches, or demonstrate to communities and funders that restoration investments are producing results.


### Community Parameter Rankings {-}

Two community workshops (June 2025) ranked prioritization parameters. The rankings diverged in instructive ways that the governance framework must hold without collapsing one perspective into the other.

```{r tab-workshop-rankings}
workshop <- tibble::tribble(
  ~Parameter,                          ~`Wet'suwet'en Rank`, ~`Wet'suwet'en Avg`, ~`Landowner Rank`, ~`Landowner Avg`,
  "Cultural Significance",             1L,                   2.16,                 4L,                4.33,
  "Probability of Success",            2L,                   2.83,                 2L,                2.50,
  "Increase Watershed Function",       3L,                   3.16,                 1L,                2.00,
  "Proximity to High-Value Fish Habitat", 4L,                3.50,                 3L,                3.00,
  "Collective Wellbeing",              5L,                   4.00,                 NA_integer_,        NA_real_,
  "Opportunity for Engagement",        6L,                   5.16,                 6L,                5.33,
  "Economic Impact",                   7L,                   6.50,                 5L,                5.00,
  "Land Ownership",                    8L,                   7.50,                 7L,                5.83
)

knitr::kable(workshop, caption = "Prioritization parameter rankings from Wet'suwet'en and landowner community workshops (June 2025). Wet'suwet'en participants ranked Cultural Significance decisively first; landowner participants ranked Watershed Function first and proposed a decision-tree approach. Both groups ranked Probability of Success highly.")
```

Both groups value Watershed Function and Fish Habitat highly. The critical difference: Wet'suwet'en participants ranked Cultural Significance decisively first (average score 2.16), while landowner participants ranked it fourth. This is not a conflict to resolve — it reflects fundamentally different relationships to the land. Wet'suwet'en participants emphasized that "when we heal the land we heal ourselves" — restoration is cultural, spiritual, and intergenerational, not just technical. Landowner participants proposed a practical decision-tree approach: Watershed Function first, then Probability of Success plus landowner willingness, then Fish Habitat, then score remaining criteria 0–5. The Collective Wellbeing parameter was added by Wet'suwet'en participants — it was not in the original parameter set.


### Addressing Cumulative Pressures {-}

Before we prioritize which projects to do, we must also consider whether cumulative pressures on floodplain function are still increasing. Land cover change detection ([Appendix - LULC](#app-lulc)) shows ongoing floodplain vegetation loss across most sub-basins — if restoration gains on one reach are offset by continued riparian clearing elsewhere, the net watershed benefit is diminished. Where losses far exceed any realistic restoration capacity — as in sub-basins showing double-digit percentage declines in floodplain tree cover over just six years — the most effective use of limited resources may be preventing further damage rather than implementing new restoration projects. This calculus is sub-basin specific: in reaches with high ongoing loss, protection and land use engagement may deliver more ecological benefit per dollar than engineered interventions, while in more stable reaches, targeted restoration can build on existing function. The boundaries of these priority zones are flexible and should be updated as new land cover data becomes available. Complementary actions to site-level restoration include supporting compliance with existing riparian setback regulations, identifying ongoing sources of channel and floodplain modification, protecting remaining wetlands and beaver complexes, and communicating the connection between floodplain land use and downstream impacts.


### Applying the Gates: Project Types That Pass {-}

The gates are practical. To illustrate how they would be applied, the following are examples of project types and locations where the available data suggests all three gates could be met. These are preliminary candidates — site-level restoration designs with detailed mapping are in development to support governance review and decision-making.

**Riparian fencing and vegetation recovery in Cesford-Ailport** — This sub-basin has the largest modelled floodplain below Bulkley Falls (1,060 ha) and the highest coho spawning length (60.8 km). Existing restoration investments provide baseline monitoring sites. The diagnostic picture is clear from land cover change analysis showing 19.9% agriculture expansion into the floodplain (G1 passes). Active vegetation removal is documented and ongoing (G2 — address pressure and restore simultaneously). With 648 ha of private floodplain and 564 ha of Crown land, there are multiple potential partners on both tenure types, and prior landowner engagement provides a foundation for access (G3). This sub-basin demonstrates how protection and restoration can work together — fencing cattle out of riparian areas on private land while restoring native vegetation on adjacent Crown parcels. Each site becomes a shared learning environment where landowners, practitioners, and Wet'suwet'en knowledge holders work side by side — building understanding of riparian function, cultural significance, and land stewardship across disciplinary boundaries.

**Process-based restoration and floodplain stewardship in Maxan Creek** — Maxan Creek has the highest agriculture expansion rate within its modelled floodplain (59.4%), with substantial floodplain vegetation removal documented through land cover change analysis since 2017. Field assessment at effectiveness monitoring site MX2 revealed evidence of dyking to keep the stream off the floodplain. The sub-basin contains 97 ha of First Nations reserve land across 3 reserves, and documented cultural sites including Tsaslachque at the Maxan Creek–Foxy Creek confluence, a historic Wet'suwet'en fishing site for sockeye, coho, chinook, and trout. Although salmon have not been observed in Maxan Creek in the modern monitoring record, the adjacent Nehl' dzee tez diee at the Bulkley Lake outlet was a fishing site for sockeye, coho, and chinook — documenting salmon use above Bulkley Falls that predates current observation networks. The extensive floodplain in this sub-basin has high potential for retaining water on the landscape and releasing it slowly to downstream reaches, helping moderate low flows and temperature stress during summer. Process-based restoration through beaver dam analogues and riparian revegetation can reconnect floodplain function without requiring landowner agreements on every parcel — work on Crown and First Nations reserve lands passes the access gate (G3) directly. The diagnostic basis is strong (G1): land cover change documents the trajectory, and field sites provide baseline data. The approach addresses root cause (G2): loss of water storage and riparian function on the floodplain. With First Nations reserve lands, documented cultural significance, and proximity to Wet'suwet'en communities, Maxan Creek is a natural setting for interdisciplinary field learning — where restoration practitioners, community members, and Wet'suwet'en knowledge holders work together on process-based restoration while reconnecting with the cultural history of the landscape.

**Monitoring and assessment in Bulkley River Above Falls** — This sub-basin has the second-highest absolute tree loss (165 ha) and 213 km of coho habitat, yet has 1 documented cultural fishing site and no reserves. Before prescribing treatments, we need to understand the pattern of loss through historic imagery analysis and field assessment. This is a G1 response — routing to the assessment queue rather than premature project design. The monitoring itself is an investment in diagnostic certainty for future prioritization rounds. The monitoring and assessment work is itself an educational opportunity — field days that bring together biologists, Wet'suwet'en knowledge holders, landowners, and students to collectively document conditions and build shared understanding of watershed health.


## Specific Recommendations {-}

The following recommendations predate the prioritization framework above and represent broad program areas — ecosystem health monitoring, range management, education, and floodplain recovery. As the framework matures, these program areas will be evaluated through the diagnostic gates and scoring principles, with site-level projects within each program scored and ranked accordingly ([#130](https://github.com/NewGraphEnvironment/restoration_wedzin_kwa_2024/issues/130)).

```{r, echo=FALSE}
knitr::include_url("https://www.newgraphenvironment.com/restoration_wedzin_kwa_2024_recomendations/", height = "800px")
```


```{r}
# 1.  Amalgamate effectiveness monitoring information and present within online reporting to facilitate adaptive management through sharing and review. Evolve program to leverage past monitoring methodology and data include an updated monitoring schedule, monitoring objectives, and methods. Methodology using Watershed Restoration Procedures - Routine effectiveness evaluation procedures and riparian inventory techniques ([LMH25](https://www.for.gov.bc.ca/hfd/pubs/Docs/Lmh/Lmh25/Lmh25_ed2_(2010).pdf) - percent cover by structural stage, species composition, etc.) could be utilized using pre-built digital forms served to field teams from the collaborative GIS project.
# 
# 1.  Design long term water monitoring program to leverage past work and attempt to quantify upstream water quality impacts via tools such as [CABIN](https://www.canada.ca/en/environment-climate-change/services/canadian-aquatic-biomonitoring-network/resources.html) sampling program [@canada2008CanadianAquatic], 5 sample in 30 day water quality sampling during both high and low flow periods [@wlrs2024BritishColumbia; @moe2024BritishColumbia], continuation/expansion of [temperature monitoring](https://public.tableau.com/app/profile/skeena.knowledge.trust/viz/UBRWaterTemperatureMonitoringDashboardDraft/UBRWaterTemp_Dashboard) [@skeenaknowledgetrustUBRWater, @westcott2022UpperBulkleya], quantify amounts of water withdrawn by licensees in the upper Bulkley catchment during the April to September low flow period, etc).  Importantly, careful consideration of how to implement water quality program recommendations from  @price2014UpperBulkley, @oliver2020Analysis2017 and @westcott2022UpperBulkleya during this process is also recommended.
# 
# 
# 2. Incentivize and help coordinate range tenure practice improvements (off-site watering, rotational field use, riparian exclusion, monitoring, range use plan updates that reduce cattle load, range stewardship planning that includes "red" zones for cattle exclusion, etc.) to reduce cattle impacts on riparian and water quality values.  Cattle waste loading to tributary and mainstem habitats is of paramount concern as water quality degradation due to high biological oxygen demand related to nutrient loading will undermine the ability of aquatic life to recover from cumulative impacts in the watershed.
# 
# 3.  Leverage and expand existing education programs for youth related to watershed health and cultural/ecosystem significance of fish/riparian/wildlife values to provide outreach for all school aged and post-secondary students in the greater Skeena region.
# 
# 4.  Work with organizations such as the Woodmere Nursery, Skeena based "Seed the North", knowledge holders and others to develop a local seed collection and propagation program to increase capacity for floodplain revegetation that can be scaled up to meet the challenges faced by historic removal of these native vegetation communities.
# 
# 10. In our recent field assessments, we observed that the protection of road and rail infrastructure through streambank armoring is not adequately incorporating best practices for vegetating riprap, soft armouring where possible and establishing/restoring effective riparian buffers. This oversight can significantly impact watershed health and biodiversity. To address this issue, we recommend collaborating with relevant stakeholders to advocate for the documentation, evolution, and dissemination of best practices related to streambank armoring.
# 
# 4.  Develop a schedule for on the ground monitoring via pre-planned instream surveys (walks) of mainstem and tributary habitats and riparian recovery/monitoring. Map these "trail" systems and document values within them. Increasing the presence of First Nation representatives, environmental practitioners and regulators within valuable streams within the Neexdzii Kwah watershed - regardless of land ownership - will help educate surveyors on the current state of these systems, provide information regarding recovery effort results and incentivize responsible land management by landowners/managers in the region.
# 
# 
# 6.  Select and refine parameters to prioritize future recovery program type (ex. water quality monitoring, range management program, education initiative, riparian vegetation recovery, expanded seed collection/propogation program, erosion protection, etc.) and allocate available dollars accordingly.  After financial and capacity resources are delineated by recovery program type - select parameters, and weight a ranking system for each to inform prioritization of specific sites within each recovery program. For physical actions such as riparian fencing for cattle exclusion, riparian/floodplain vegetation recovery and erosion protection include prioritization criteria that incorporates the amount of riparian/floodplain area available for recovery (ie. the wider the buffer available for meaningful native floodplain vegetation community establishment -  the higher the priority for investment).  Utilize established standards when possible such as the [Riparian Areas Protection Regulation](https://www2.gov.bc.ca/gov/content/environment/plants-animals-ecosystems/fish/aquatic-habitat-management/riparian-areas-regulation) and the [Riparian Management Area Guidebook](https://www2.gov.bc.ca/gov/content/industry/forestry/managing-our-forest-resources/silviculture/silvicultural-systems/silviculture-guidebooks/riparian-management-area-guidebook) [@ministryofforests1995RiparianManagement].
# 
# 7. Working with watershed users—including industrial players like railways and highways, as well as landowners and range tenure holders—is essential for promoting water retention through beaver activity. The Beaver Restoration Assessment Tool (BRAT) can guide efforts by identifying areas where beaver activity can optimize water retention and flow regulation. Beavers are natural ecosystem engineers; their dams slow water flow, enhance groundwater recharge, and reduce flood risks, while also improving water quality by trapping sediments and filtering pollutants. Providing incentives to landowners and tenure holders to support water retention on the landscape can help catalyze beaver influenced wetlands throughout the watershed addressing potential land use conflicts while aligning ecological goals with economic interests.
# 
# 5. Acquire, georeference, archive (within SKT) and analyze historic aerial imagery for the Neexdzii Kwah watershed and compare with recent orthoimagery and/or LiDAR data to quantify historic changes in stream morphology and floodplain extent resulting in loss of quantity and quality of water and fish habitat while highlighting areas of historic dredging, realignment, and floodplain disconnections due to infrastructure, to guide future restoration efforts. Script all analysis when possible to facilitate reproducibility and sharing of the analysis.
# 
# 8.  Continue to develop and document data management workflows to leverage existing established data storage systems to retrieve data from and load to and build capacity for all interested to do the same. Established data storage systems include the [BC Data Catalogue](https://catalogue.data.gov.bc.ca/), the [Skeena Salmon Data Centre](https://data.skeenasalmon.info/), [NuSEDS-New Salmon Escapement Database System](https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6) and [bcfishpass](https://github.com/smnorris/bcfishpass/tree/main/data)
# 
# 
# 9.  Amalgamate drone acquired ortho-imagery gathered for sites in the Neexdzii Kwah watershed, store as Cloud Optimized Geotiffs on a cloud service provider and link to the imagery via the collaborative GIS project. Orthoimagery has been gathered for past monitoring of historic restoration sites as well as part of fish passage restoration planning efforts.
# 
# 
# 
# 
# 
# 12. Link historic and potential future sites to the freshwater atlas using the linear feature id of stream segments and conduct watershed drainage area analysis for individual stream segments (or portions of steram segments) so that this area can be tied to other GIS and user input parameters. Functions kept in [fwapg](https://github.com/smnorris/fwapg), [fwapgr](https://github.com/poissonconsulting/fwapgr), [fwatlasbc](https://github.com/poissonconsulting/fwatlasbc) and custom function developed for this project can be used. Results from this can be compared to other methods for determining project setbacks from the high water mark such as the [Riparian Areas Protection Regulation](https://www2.gov.bc.ca/gov/content/environment/plants-animals-ecosystems/fish/aquatic-habitat-management/riparian-areas-regulation) and the [Riparian Management Area Guidebook](https://www2.gov.bc.ca/gov/content/industry/forestry/managing-our-forest-resources/silviculture/silvicultural-systems/silviculture-guidebooks/riparian-management-area-guidebook) [@ministryofforests1995RiparianManagement].
# 
# 
# 14.  Spatially represent historic fisheries sites presented in Table \@ref(tab:tab-hist-sites) so the option is available that potential areas of restoration can be ranked according to their proximity to these sites.
# 
# 15.  Spatially delineate areas of high fisheries values based on known spawning and rearing habitat for fish species of interest. Where data gaps exist - develop study plans to understand where this high value habitat is and quantify its value for fish species of interest through standardized methodologies such as Fish and Fish Habitat Assessment Procedures [@johnstonFishHabitatAssessment1996]. For known spawning and rearing locations, load reviewed data into `bcfishpass` using the `user_habitat_classification.csv` file located [here](https://github.com/smnorris/bcfishpass/blob/main/data/user_habitat_classification.csv) so it can be pulled into the collaborative GIS project and queried to facilitate priority ranking of areas of high fisheries values.
```

