# **Appendix 1 - Climate Anomaly Data** {-#app-climate-anomaly}

```{r setup-app-climate-anomaly}
knitr::opts_chunk$set(fig.path = "fig/app-climate-anomaly/", dev = "png")
```

Climate anomaly data for the Neexdzii Kwah study area were extracted from ERA5-Land reanalysis (1951–`r max(readr::read_csv(here::here("data", "climate", "neexdzii_kwah_anomaly_timeseries.csv"), show_col_types = FALSE)$yr)`) and analyzed for seven climate parameters across seasonal and annual periods. Three key patterns emerge:

- **The watershed is roughly 2°C warmer than it was in the mid-20th century.** At a rate of approximately 0.03°C per year, this warming has accumulated over seven decades to produce a substantial shift. Comparing the last decade to the pre-1980 period, summers are about 2°C warmer. Since 2013, nearly every year has been above the 1981–2010 average. These trends are highly statistically significant across all seasons (p < 0.001).
- **Precipitation has not changed.** No season shows a statistically significant trend in total precipitation, though year-to-year variability is high.
- **Soils are drying despite stable precipitation.** Summer soil moisture has shifted by roughly 5 percentage points relative to the pre-1980 period and has been below normal in most years since 2000. Warmer temperatures drive more evapotranspiration — pulling moisture from soils even when the same amount of rain falls. For fish, drier soils mean reduced summer baseflows during the period when flows are already at their lowest.

```{r tab-climate-trends-cap, results="asis"}
my_caption <- "Climate anomaly trend statistics for the Neexdzii Kwah study area (ERA5-Land reanalysis). Slope is the rate of change per year (Theil-Sen estimator). Total Change is the slope multiplied by the number of years in the period. p-values below 0.05 indicate statistically significant trends (Mann-Kendall test). Two trend periods are shown: 1950–present captures the full record, while 1980–present isolates more recent change."
my_tab_caption(tip_flag = FALSE)
```

```{r tab-climate-trends}
trends <- readr::read_csv(
  here::here("data", "climate", "neexdzii_kwah_anomaly_trends.csv"),
  show_col_types = FALSE
)

unit_lookup <- c(
  tmean = "\u00b0C", tmax = "\u00b0C", tmin = "\u00b0C",
  prcp = "%", vpd = "Pa", rh = "%", soil_moisture = "%"
)

par_labels <- c(
  tmean = "Mean temp.", tmax = "Max temp.", tmin = "Min temp.",
  prcp = "Precipitation", vpd = "VPD", rh = "Relative humidity",
  soil_moisture = "Soil moisture"
)

trends_display <- trends |>
  dplyr::filter(period %in% c("annual", "winter", "spring", "summer", "fall")) |>
  dplyr::mutate(
    Parameter = par_labels[par],
    Period = stringr::str_to_title(period),
    Slope = round(trend_slope, 3),
    Years = n_years,
    `Total Change` = round(trend_slope * n_years, 1),
    Unit = unit_lookup[par],
    `p-value` = mk_p
  ) |>
  dplyr::select(Parameter, Period, Slope, Years, `Total Change`, Unit, `p-value`) |>
  dplyr::arrange(Parameter, Period, Years)

trends_display |>
  my_dt_table(page_length = 20, cols_freeze_left = 0)
```

<br>

Spatial anomaly maps and time series trend plots below were generated using the [BC Climate Anomaly app](https://bcgov-env.shinyapps.io/bc_climate_anomaly/) developed by the BC Ministry of Forests Future Forest Ecosystems Centre. Anomalies are relative to the 1981–2010 climatological normal. Trend lines show Theil-Sen slopes for 1950–present (black dashed) and 1980–present (blue solid), with Mann-Kendall p-values.

```{r plot-climate-anomaly, echo=FALSE, results='asis'}
files <- fs::dir_ls("data/inputs_raw/climate_anomaly", glob = "*.png")
for (f in files) {
  cat("![](", f, ")\n\n")
}
```
