# **Appendix - Land Use Land Cover Change** {-#app-lulc}

```{r setup-app-lulc}
knitr::opts_chunk$set(fig.path = "fig/app-lulc/", dev = "png")
```

```{r lulc-load, message=FALSE, warning=FALSE}
library(drift)
library(sf)
library(terra)
library(ggplot2)
library(dplyr)
library(tidyr)

sf::sf_use_s2(FALSE)
terra::terraOptions(threads = 12)

# Sub-basins from watershed picker
subbasins <- sf::st_read(
  here::here("data", "lulc", "subbasins.gpkg"),
  quiet = TRUE
)

# Floodplain AOI
floodplain <- sf::st_read(
  here::here("data", "lulc", "floodplain_neexdzii_co.gpkg"),
  quiet = TRUE
)

# Label sub-basins: auto-number by stream name, 1 = mouth (lowest DRM)
subbasins <- subbasins |>
  dplyr::arrange(gnis_name, drm) |>
  dplyr::mutate(
    stream_short = sub(" (Creek|River|Lake)$", "", gnis_name),
    seq = ave(seq_len(dplyr::n()), gnis_name, FUN = seq_along),
    label = paste(stream_short, seq)
  )
```

## Floodplain Overview

The floodplain area of interest covers `r round(as.numeric(sf::st_area(sf::st_union(floodplain))) / 1e6, 1)` km² of valley bottom along the Neexdzii Kwa (Bulkley River) and tributaries, delineated using the Valley Confinement Algorithm via the `flooded` package.

```{r map-lulc-subbasins, fig.cap="Sub-basin boundaries used for land cover comparison.", fig.width=10, fig.height=12}
cache <- here::here("data", "spatial")
streams_all <- readRDS(file.path(cache, "streams.rds"))
roads    <- readRDS(file.path(cache, "roads.rds"))
railway  <- readRDS(file.path(cache, "railway.rds"))

towns <- sf::st_sf(
  name = c("Houston", "Smithers", "Burns Lake", "Topley"),
  geometry = sf::st_sfc(
    sf::st_point(c(-126.648, 54.398)),
    sf::st_point(c(-127.176, 54.779)),
    sf::st_point(c(-125.764, 54.230)),
    sf::st_point(c(-126.246, 54.566)),
    crs = 4326
  )
)

highways <- roads |> dplyr::filter(road_type == "RH1")

# Sub-basin label coordinates
sb_pts <- sf::st_point_on_surface(subbasins)
sb_label_df <- data.frame(
  label = subbasins$label,
  lon = sf::st_coordinates(sb_pts)[, 1],
  lat = sf::st_coordinates(sb_pts)[, 2]
)

map_bbox <- sf::st_bbox(subbasins)

ggplot() +
  geom_sf(data = subbasins, aes(fill = label), alpha = 0.25, color = "grey40",
          linewidth = 0.5, show.legend = FALSE) +
  geom_sf(data = floodplain, fill = "darkgreen", color = "darkgreen",
          alpha = 0.3, linewidth = 0.6) +
  geom_sf(data = streams_all |> dplyr::filter(stream_order >= 5),
          color = "#7ba7cc", linewidth = 0.4) +
  geom_sf(data = railway, color = "grey30", linewidth = 0.6, linetype = "dashed") +
  geom_sf(data = highways, color = "#c0392b", linewidth = 0.8) +
  geom_sf(data = towns, size = 2.5, shape = 16) +
  geom_label(data = sb_label_df, aes(x = lon, y = lat, label = label),
    size = 3, fontface = "bold",
    fill = "white", alpha = 0.8,
    label.padding = unit(0.15, "lines")
  ) +
  geom_text(data = as.data.frame(sf::st_coordinates(towns)) |>
    dplyr::mutate(name = towns$name),
    aes(x = X, y = Y + 0.02, label = name),
    size = 3, fontface = "bold") +
  coord_sf(xlim = c(map_bbox["xmin"] - 0.04, map_bbox["xmax"] + 0.04),
           ylim = c(map_bbox["ymin"] - 0.06, map_bbox["ymax"] + 0.06)) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = -Inf,
           fill = "darkgreen", alpha = 0) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank()
  )
```

## Land Cover by Sub-Basin

IO Land Use Land Cover (10 m) is fetched from Planetary Computer for 2017, 2020, and 2023 and classified using the `drift` pipeline. Each sub-basin's floodplain is analyzed independently.

```{r lulc-classify, message=FALSE, warning=FALSE, cache=TRUE}
# Clip floodplain by each sub-basin and run drift pipeline
results <- list()

for (i in seq_len(nrow(subbasins))) {
  sb <- subbasins[i, ]
  fp_clip <- sf::st_intersection(floodplain, sb) |>
    sf::st_collection_extract("POLYGON") |>
    sf::st_union() |>
    sf::st_sf(geometry = _)
  sf::st_crs(fp_clip) <- sf::st_crs(floodplain)

  if (nrow(fp_clip) == 0 || as.numeric(sf::st_area(fp_clip)) < 1e4) {
    message("Skipping ", sb$label, " — no floodplain overlap")
    next
  }

  message("Processing: ", sb$label)
  rasters <- dft_stac_fetch(fp_clip, source = "io-lulc", years = c(2017, 2020, 2023))
  classified <- dft_rast_classify(rasters, source = "io-lulc")
  summary <- dft_rast_summarize(classified, unit = "ha")
  summary$subbasin <- sb$label
  results[[i]] <- summary
}

lulc_all <- bind_rows(results)
```

```{r lulc-prep}
# Prepare wide table and transition summaries used in multiple sections
wide <- lulc_all |>
  select(subbasin, year, class_name, pct) |>
  pivot_wider(names_from = year, values_from = pct) |>
  arrange(subbasin, class_name)

# Compute change from 2017 to 2023 for each class and sub-basin
lulc_delta <- lulc_all |>
  filter(year %in% c(2017, 2023)) |>
  select(subbasin, year, class_name, pct) |>
  pivot_wider(names_from = year, values_from = pct, names_prefix = "y") |>
  mutate(delta = y2023 - y2017) |>
  arrange(subbasin, class_name)
```

```{r plot-lulc-change, fig.cap="Land cover change by sub-basin (2017-2023).", fig.width=12, fig.height=8}
# Focus on major classes
major <- lulc_all |>
  group_by(class_name) |>
  summarize(max_pct = max(pct), .groups = "drop") |>
  filter(max_pct >= 5) |>
  pull(class_name)

lulc_all |>
  filter(class_name %in% major) |>
  mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = pct, fill = class_name)) +
  geom_col(position = "dodge") +
  facet_wrap(~subbasin, scales = "free_y") +
  scale_fill_brewer(palette = "Set2", name = "Land Cover") +
  theme_minimal() +
  labs(
    y = "% of Floodplain",
    x = NULL,
    title = "Land Cover Change by Sub-Basin (IO LULC 10m)"
  )
```

## Trees vs Rangeland

The conversion of Trees to Rangeland is a key indicator of riparian disturbance.

```{r plot-lulc-trees-range, fig.cap="Trees and Rangeland proportion by sub-basin over time.", fig.width=12, fig.height=6}
lulc_all |>
  filter(class_name %in% c("Trees", "Rangeland")) |>
  mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = pct, color = class_name, group = class_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~subbasin, nrow = 2) +
  scale_color_manual(values = c("Trees" = "#006400", "Rangeland" = "#FFBB22"), name = NULL) +
  theme_minimal() +
  labs(
    y = "% of Floodplain",
    x = NULL,
    title = "Trees vs Rangeland by Sub-Basin"
  )
```

## Sub-Basin Ranking by Tree Loss

Sub-basins ranked by the change in Trees and Rangeland cover between 2017 and 2023. Negative tree delta indicates loss; positive rangeland delta indicates gain.

```{r plot-lulc-ranking, fig.cap="Sub-basins ranked by tree cover loss (2017-2023).", fig.width=10, fig.height=6}
ranking <- lulc_delta |>
  filter(class_name %in% c("Trees", "Rangeland")) |>
  select(subbasin, class_name, delta) |>
  pivot_wider(names_from = class_name, values_from = delta) |>
  mutate(net_shift = Rangeland - Trees) |>
  arrange(desc(net_shift))

ranking_long <- ranking |>
  pivot_longer(cols = c(Trees, Rangeland), names_to = "class_name", values_to = "delta")

ranking_long |>
  mutate(subbasin = factor(subbasin, levels = ranking$subbasin)) |>
  ggplot(aes(x = subbasin, y = delta, fill = class_name)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Trees" = "#006400", "Rangeland" = "#FFBB22"), name = NULL) +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    y = "Change in % of Floodplain (2017-2023)",
    x = NULL,
    title = "Sub-Basins Ranked by Tree-to-Rangeland Conversion"
  )
```

## Class Transitions

To understand whether tree loss is primarily converting to rangeland or also shifting to other classes (e.g., Built Area, Crops), we examine all class changes across the study period.

```{r plot-lulc-all-delta, fig.cap="Change in land cover by class and sub-basin (2017-2023).", fig.width=12, fig.height=8}
lulc_delta |>
  filter(abs(delta) >= 0.5) |>
  mutate(subbasin = factor(subbasin)) |>
  ggplot(aes(x = reorder(class_name, delta), y = delta, fill = delta > 0)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~subbasin, scales = "free_x") +
  scale_fill_manual(values = c("TRUE" = "#FFBB22", "FALSE" = "#006400")) +
  coord_flip() +
  theme_minimal() +
  labs(
    y = "Change in % of Floodplain (2017-2023)",
    x = NULL,
    title = "All Land Cover Class Changes by Sub-Basin"
  )
```

```{r lulc-transition-summary}
# Aggregate across all sub-basins to see overall class shifts
overall_delta <- lulc_delta |>
  group_by(class_name) |>
  summarize(
    mean_delta = round(mean(delta, na.rm = TRUE), 2),
    min_delta = round(min(delta, na.rm = TRUE), 2),
    max_delta = round(max(delta, na.rm = TRUE), 2),
    n_gaining = sum(delta > 0.5, na.rm = TRUE),
    n_losing = sum(delta < -0.5, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(mean_delta)
```

```{r tab-lulc-transitions-cap, results="asis", eval=gitbook_on}
my_caption <- "Mean land cover change (% of floodplain, 2017-2023) across all sub-basins. Negative values indicate loss."

my_tab_caption()
```

```{r tab-lulc-transitions, eval=gitbook_on}
overall_delta |>
  my_dt_table(cols_freeze_left = 1, page_length = 15)
```

```{r tab-lulc-transitions-static, eval=identical(gitbook_on, FALSE)}
knitr::kable(overall_delta, digits = 2,
  caption = "Mean land cover change (% of floodplain, 2017-2023) across all sub-basins. Negative values indicate loss.")
```

## Land Cover Composition

```{r tab-lulc-wide-cap, results="asis", eval=gitbook_on}
my_caption <- "Land cover composition (% of floodplain) by sub-basin and year."

my_tab_caption()
```

```{r tab-lulc-wide, eval=gitbook_on}
wide |>
  my_dt_table(cols_freeze_left = 2, page_length = 25)
```

```{r tab-lulc-wide-static, eval=identical(gitbook_on, FALSE)}
knitr::kable(wide, digits = 1,
  caption = "Land cover composition (% of floodplain) by sub-basin and year.")
```
