# **Appendix - Land Use Land Cover Change** {-#app-lulc}

```{r setup-app-lulc}
knitr::opts_chunk$set(fig.path = "fig/app-lulc/", dev = "png")
```

```{r lulc-load, message=FALSE, warning=FALSE}
library(drift)
library(sf)
library(terra)
library(ggplot2)
library(dplyr)
library(tidyr)

sf::sf_use_s2(FALSE)
terra::terraOptions(threads = 12)

# Sub-basins from watershed picker
subbasins <- sf::st_read(
  here::here("data", "lulc", "subbasins.gpkg"),
  quiet = TRUE
)

# Floodplain AOI
floodplain <- sf::st_read(
  here::here("data", "lulc", "floodplain_neexdzii_co.gpkg"),
  quiet = TRUE
)

# Label sub-basins: auto-number by stream name, 1 = mouth (lowest DRM)
subbasins <- subbasins |>
  dplyr::arrange(gnis_name, drm) |>
  dplyr::mutate(
    stream_short = sub(" (Creek|River|Lake)$", "", gnis_name),
    seq = ave(seq_len(dplyr::n()), gnis_name, FUN = seq_along),
    label = paste(stream_short, seq)
  )
```

## Floodplain Overview

The floodplain area of interest covers `r round(as.numeric(sf::st_area(sf::st_union(floodplain))) / 1e6, 1)` km² of valley bottom along the Neexdzii Kwa (Bulkley River) and tributaries, delineated using the Valley Confinement Algorithm via the `flooded` package.

```{r map-lulc-subbasins, fig.cap="Sub-basin boundaries used for land cover comparison.", fig.width=10, fig.height=12}
cache <- here::here("data", "spatial")
streams_all <- readRDS(file.path(cache, "streams.rds"))
roads    <- readRDS(file.path(cache, "roads.rds"))
railway  <- readRDS(file.path(cache, "railway.rds"))

towns <- sf::st_sf(
  name = c("Houston", "Smithers", "Burns Lake", "Topley"),
  geometry = sf::st_sfc(
    sf::st_point(c(-126.648, 54.398)),
    sf::st_point(c(-127.176, 54.779)),
    sf::st_point(c(-125.764, 54.230)),
    sf::st_point(c(-126.246, 54.566)),
    crs = 4326
  )
)

highways <- roads |> dplyr::filter(road_type == "RH1")

# Sub-basin label points
sb_labels <- subbasins |>
  dplyr::mutate(geometry = sf::st_point_on_surface(geom)) |>
  sf::st_set_geometry("geometry")

map_bbox <- sf::st_bbox(subbasins)

ggplot() +
  geom_sf(data = subbasins, aes(fill = label), alpha = 0.25, color = "grey40",
          linewidth = 0.5, show.legend = FALSE) +
  geom_sf(data = floodplain, aes(color = "Floodplain"), fill = "darkgreen", alpha = 0.3,
          linewidth = 0.6, show.legend = "polygon") +
  geom_sf(data = streams_all |> dplyr::filter(stream_order >= 5),
          color = "#7ba7cc", linewidth = 0.4) +
  geom_sf(data = railway, aes(linetype = "Railway"), color = "grey30",
          linewidth = 0.6, show.legend = "line") +
  geom_sf(data = highways, aes(linetype = "Highway"), color = "#c0392b",
          linewidth = 0.8, show.legend = "line") +
  geom_sf(data = towns, size = 2.5, shape = 16) +
  ggrepel::geom_label_repel(
    data = sb_labels,
    aes(label = label, geometry = geometry),
    stat = "sf_coordinates",
    size = 3, fontface = "bold",
    fill = "white", alpha = 0.8,
    label.padding = unit(0.15, "lines"),
    min.segment.length = 0,
    max.overlaps = 20
  ) +
  geom_sf_text(data = towns, aes(label = name), size = 3, fontface = "bold",
               nudge_y = 0.02) +
  coord_sf(xlim = c(map_bbox["xmin"] - 0.04, map_bbox["xmax"] + 0.04),
           ylim = c(map_bbox["ymin"] - 0.06, map_bbox["ymax"] + 0.06)) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_color_manual(name = NULL, values = c("Floodplain" = "darkgreen"),
                     guide = guide_legend(order = 1, override.aes = list(fill = "darkgreen", alpha = 0.3))) +
  scale_linetype_manual(name = NULL, values = c("Highway" = "solid", "Railway" = "dashed"),
                        guide = guide_legend(order = 2,
                          override.aes = list(color = c("#c0392b", "grey30")))) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(
    legend.position = c(0.85, 0.15),
    legend.background = element_rect(fill = "white", color = "grey80"),
    legend.key.size = unit(0.8, "lines"),
    panel.grid.minor = element_blank()
  )
```

## Land Cover by Sub-Basin

IO Land Use Land Cover (10 m) is fetched from Planetary Computer for 2017, 2020, and 2023 and classified using the `drift` pipeline. Each sub-basin's floodplain is analyzed independently.

```{r lulc-classify, message=FALSE, warning=FALSE, cache=TRUE}
# Clip floodplain by each sub-basin and run drift pipeline
results <- list()

for (i in seq_len(nrow(subbasins))) {
  sb <- subbasins[i, ]
  fp_clip <- sf::st_intersection(floodplain, sb) |>
    sf::st_collection_extract("POLYGON") |>
    sf::st_union() |>
    sf::st_sf(geometry = _)
  sf::st_crs(fp_clip) <- sf::st_crs(floodplain)

  if (nrow(fp_clip) == 0 || as.numeric(sf::st_area(fp_clip)) < 1e4) {
    message("Skipping ", sb$label, " — no floodplain overlap")
    next
  }

  message("Processing: ", sb$label)
  rasters <- dft_stac_fetch(fp_clip, source = "io-lulc", years = c(2017, 2020, 2023))
  classified <- dft_rast_classify(rasters, source = "io-lulc")
  summary <- dft_rast_summarize(classified, unit = "ha")
  summary$subbasin <- sb$label
  results[[i]] <- summary
}

lulc_all <- bind_rows(results)
```

```{r tab-lulc-wide, results='asis'}
wide <- lulc_all |>
  select(subbasin, year, class_name, pct) |>
  pivot_wider(names_from = year, values_from = pct) |>
  arrange(subbasin, class_name)

knitr::kable(wide, digits = 1,
  caption = "Land cover composition (% of floodplain) by sub-basin and year.")
```

```{r plot-lulc-change, fig.cap="Land cover change by sub-basin (2017-2023).", fig.width=12, fig.height=8}
# Focus on major classes
major <- lulc_all |>
  group_by(class_name) |>
  summarize(max_pct = max(pct), .groups = "drop") |>
  filter(max_pct >= 5) |>
  pull(class_name)

lulc_all |>
  filter(class_name %in% major) |>
  mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = pct, fill = class_name)) +
  geom_col(position = "dodge") +
  facet_wrap(~subbasin, scales = "free_y") +
  scale_fill_brewer(palette = "Set2", name = "Land Cover") +
  theme_minimal() +
  labs(
    y = "% of Floodplain",
    x = NULL,
    title = "Land Cover Change by Sub-Basin (IO LULC 10m)"
  )
```

## Trees vs Rangeland

The conversion of Trees to Rangeland is a key indicator of riparian disturbance.

```{r plot-lulc-trees-range, fig.cap="Trees and Rangeland proportion by sub-basin over time.", fig.width=12, fig.height=6}
lulc_all |>
  filter(class_name %in% c("Trees", "Rangeland")) |>
  mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = pct, color = class_name, group = class_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~subbasin, nrow = 2) +
  scale_color_manual(values = c("Trees" = "#006400", "Rangeland" = "#FFBB22"), name = NULL) +
  theme_minimal() +
  labs(
    y = "% of Floodplain",
    x = NULL,
    title = "Trees vs Rangeland by Sub-Basin"
  )
```
