# **Appendix - Land Use Land Cover Change** {-#app-lulc}

```{r setup-app-lulc}
knitr::opts_chunk$set(fig.path = "fig/app-lulc/", dev = "png")
```

```{r lulc-load, message=FALSE, warning=FALSE}
library(drift)
library(sf)
library(terra)
library(tmap)
library(maptiles)
library(ggplot2)
library(dplyr)
library(tidyr)

sf::sf_use_s2(FALSE)
terra::terraOptions(threads = 12)

# Sub-basins from watershed picker (transform to 4326 to match other layers)
subbasins <- sf::st_read(
  here::here("data", "lulc", "subbasins.gpkg"),
  quiet = TRUE
) |> sf::st_transform(4326)

# Floodplain AOI
floodplain <- sf::st_read(
  here::here("data", "lulc", "floodplain_neexdzii_co.gpkg"),
  quiet = TRUE
) |> sf::st_transform(4326)

# Label sub-basins: auto-number by stream name, 1 = mouth (lowest DRM)
subbasins <- subbasins |>
  dplyr::arrange(gnis_name, drm) |>
  dplyr::mutate(
    stream_short = sub(" (Creek|River|Lake)$", "", gnis_name),
    seq = ave(seq_len(dplyr::n()), gnis_name, FUN = seq_along),
    label = paste(stream_short, seq)
  )
```

## Floodplain Overview

The floodplain area of interest covers `r round(as.numeric(sf::st_area(sf::st_union(floodplain))) / 1e6, 1)` km² of valley bottom along the Neexdzii Kwa (Bulkley River) and tributaries, delineated using the Valley Confinement Algorithm via the `flooded` package.

```{r map-lulc-subbasins, fig.cap="Sub-basin boundaries used for land cover comparison.", fig.width=8.5, fig.height=11}
cache <- here::here("data", "spatial")
streams_all <- readRDS(file.path(cache, "streams.rds"))
lakes    <- readRDS(file.path(cache, "lakes.rds"))
roads    <- readRDS(file.path(cache, "roads.rds"))
railway  <- readRDS(file.path(cache, "railway.rds"))

towns <- sf::st_sf(
  name = c("Houston", "Smithers", "Burns Lake", "Topley"),
  geometry = sf::st_sfc(
    sf::st_point(c(-126.648, 54.398)),
    sf::st_point(c(-127.176, 54.779)),
    sf::st_point(c(-125.764, 54.230)),
    sf::st_point(c(-126.246, 54.566)),
    crs = 4326
  )
)

# Dissolved roads by route
roads_dissolved <- roads |>
  dplyr::filter(!is.na(route)) |>
  dplyr::group_by(route) |>
  dplyr::summarise(geom = sf::st_union(geom), .groups = "drop")

# Bounding box from sub-basins with padding tuned for 8.5x11
bbox <- sf::st_bbox(subbasins)
bbox["ymax"] <- bbox["ymax"] + 0.06
bbox["ymin"] <- bbox["ymin"] - 0.06
bbox["xmin"] <- bbox["xmin"] - 0.04
bbox["xmax"] <- bbox["xmax"] + 0.04

# Clip streams and lakes to bbox
bbox_clip_sf <- sf::st_as_sfc(bbox) |> sf::st_set_crs(4326)
streams_clip <- streams_all |> sf::st_make_valid() |> sf::st_crop(bbox_clip_sf)
lakes_clip   <- lakes |> sf::st_make_valid() |> sf::st_crop(bbox_clip_sf)

# Stream labels (5th order+ named)
stream_label_pts <- streams_clip |>
  dplyr::filter(!is.na(gnis_name) & stream_order >= 5) |>
  dplyr::group_by(gnis_name) |>
  dplyr::summarise(geom = sf::st_union(geom), .groups = "drop") |>
  dplyr::mutate(geometry = sf::st_point_on_surface(geom)) |>
  sf::st_set_geometry("geometry") |>
  sf::st_set_crs(4326)

# Lake labels (>1 km²)
lake_labels <- lakes_clip[!is.na(lakes_clip$name) & lakes_clip$area_km2 > 1, ]

# Hillshade basemap
bbox_sf <- sf::st_as_sfc(bbox) |> sf::st_set_crs(4326)
relief  <- maptiles::get_tiles(bbox_sf, provider = "Esri.WorldShadedRelief",
                               zoom = 10, crop = TRUE)
basemap_stars <- stars::st_as_stars(relief)

# Earthy sub-basin palette
earth_pal <- c("#A67C52", "#C4A882", "#8B9E6B", "#D4A76A",
               "#7A8B6E", "#BFA07A", "#9B8E6E", "#C2B280")

tmap::tmap_mode("plot")

tmap::tm_shape(basemap_stars, bbox = bbox) +
  tmap::tm_rgb() +

  # Sub-basins
  tmap::tm_shape(subbasins) +
  tmap::tm_polygons(
    fill = "label",
    fill.scale = tmap::tm_scale_categorical(values = earth_pal),
    fill_alpha = 0.35,
    col = "#5C4033", lwd = 1.5,
    fill.legend = tmap::tm_legend(show = FALSE)
  ) +
  tmap::tm_text("label", size = 0.65, col = "#3B2716", fontface = "bold",
                options = tmap::opt_tm_text(shadow = TRUE)) +

  # Floodplain
  tmap::tm_shape(floodplain) +
  tmap::tm_polygons(fill = "#2d6a2e", fill_alpha = 0.30,
                    col = "#2d6a2e", lwd = 0.8) +

  # Lakes
  tmap::tm_shape(lakes_clip) +
  tmap::tm_polygons(fill = "#c6ddf0", col = "#7ba7cc", lwd = 0.4, fill_alpha = 0.85) +
  tmap::tm_shape(lake_labels) +
  tmap::tm_text("name", size = 0.55, col = "#1a5276", fontface = "italic") +

  # Streams
  tmap::tm_shape(streams_clip |> dplyr::filter(stream_order >= 4)) +
  tmap::tm_lines(col = "#7ba7cc", lwd = 0.4) +
  tmap::tm_shape(streams_clip |> dplyr::filter(stream_order >= 5)) +
  tmap::tm_lines(col = "#7ba7cc", lwd = 0.8) +
  tmap::tm_shape(stream_label_pts) +
  tmap::tm_text("gnis_name", size = 0.60, col = "#1a5276", fontface = "italic") +

  # Railway
  tmap::tm_shape(railway) +
  tmap::tm_lines(col = "black", lwd = 1.2) +
  tmap::tm_shape(railway) +
  tmap::tm_lines(col = "white", lwd = 0.6, lty = "42") +

  # Roads
  tmap::tm_shape(roads_dissolved |> dplyr::filter(route == "16")) +
  tmap::tm_lines(col = "#c0392b", lwd = 2.0) +
  tmap::tm_shape(roads_dissolved |> dplyr::filter(route != "16")) +
  tmap::tm_lines(col = "#e67e22", lwd = 1.2) +

  # Towns
  tmap::tm_shape(towns) +
  tmap::tm_dots(fill = "black", size = 0.30) +
  tmap::tm_text("name", size = 0.65, xmod = 0.8, ymod = -0.6,
                col = "grey10", fontface = "bold") +

  # Scale bar
  tmap::tm_scalebar(breaks = c(0, 10, 20),
                    position = c("left", "bottom"),
                    text.size = 0.6) +

  # Legend
  tmap::tm_add_legend(
    type = "polygons",
    labels = "Floodplain AOI",
    fill   = "#2d6a2e",
    col    = "#2d6a2e",
    lwd    = 1
  ) +
  tmap::tm_add_legend(
    type = "lines",
    labels = c("Highway 16", "Secondary highway", "Railway (CN)"),
    col  = c("#c0392b", "#e67e22", "black"),
    lwd  = c(2, 1.2, 1.2)
  ) +

  tmap::tm_layout(
    frame = TRUE,
    inner.margins  = c(0.005, 0.005, 0.005, 0.005),
    outer.margins  = c(0.002, 0.002, 0.002, 0.002),
    legend.position = c("right", "top"),
    legend.frame    = TRUE,
    legend.bg.color = "white",
    legend.bg.alpha = 0.85
  )
```

## Land Cover by Sub-Basin

IO Land Use Land Cover (10 m) is fetched from Planetary Computer for 2017, 2020, and 2023 and classified using the `drift` pipeline. Each sub-basin's floodplain is analyzed independently.

```{r lulc-classify, message=FALSE, warning=FALSE, cache=TRUE}
# Clip floodplain by each sub-basin and run drift pipeline
results <- list()

for (i in seq_len(nrow(subbasins))) {
  sb <- subbasins[i, ]
  fp_clip <- sf::st_intersection(floodplain, sb) |>
    sf::st_collection_extract("POLYGON") |>
    sf::st_union() |>
    sf::st_sf(geometry = _)
  sf::st_crs(fp_clip) <- sf::st_crs(floodplain)

  if (nrow(fp_clip) == 0 || as.numeric(sf::st_area(fp_clip)) < 1e4) {
    message("Skipping ", sb$label, " — no floodplain overlap")
    next
  }

  message("Processing: ", sb$label)
  rasters <- dft_stac_fetch(fp_clip, source = "io-lulc", years = c(2017, 2020, 2023))
  classified <- dft_rast_classify(rasters, source = "io-lulc")
  summary <- dft_rast_summarize(classified, unit = "ha")
  summary$subbasin <- sb$label
  results[[i]] <- summary
}

lulc_all <- bind_rows(results)
```

```{r lulc-prep}
# Prepare wide table and transition summaries used in multiple sections
wide <- lulc_all |>
  select(subbasin, year, class_name, pct) |>
  pivot_wider(names_from = year, values_from = pct) |>
  arrange(subbasin, class_name)

# Compute change from 2017 to 2023 for each class and sub-basin
lulc_delta <- lulc_all |>
  filter(year %in% c(2017, 2023)) |>
  select(subbasin, year, class_name, pct) |>
  pivot_wider(names_from = year, values_from = pct, names_prefix = "y") |>
  mutate(delta = y2023 - y2017) |>
  arrange(subbasin, class_name)
```

```{r plot-lulc-change, fig.cap="Land cover change by sub-basin (2017-2023).", fig.width=12, fig.height=8}
# Focus on major classes
major <- lulc_all |>
  group_by(class_name) |>
  summarize(max_pct = max(pct), .groups = "drop") |>
  filter(max_pct >= 5) |>
  pull(class_name)

lulc_all |>
  filter(class_name %in% major) |>
  mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = pct, fill = class_name)) +
  geom_col(position = "dodge") +
  facet_wrap(~subbasin, scales = "free_y") +
  scale_fill_brewer(palette = "Set2", name = "Land Cover") +
  theme_minimal() +
  labs(
    y = "% of Floodplain",
    x = NULL,
    title = "Land Cover Change by Sub-Basin (IO LULC 10m)"
  )
```

## Trees vs Rangeland

The conversion of Trees to Rangeland is a key indicator of riparian disturbance.

```{r plot-lulc-trees-range, fig.cap="Trees and Rangeland proportion by sub-basin over time.", fig.width=12, fig.height=6}
lulc_all |>
  filter(class_name %in% c("Trees", "Rangeland")) |>
  mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = pct, color = class_name, group = class_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~subbasin, nrow = 2) +
  scale_color_manual(values = c("Trees" = "#006400", "Rangeland" = "#FFBB22"), name = NULL) +
  theme_minimal() +
  labs(
    y = "% of Floodplain",
    x = NULL,
    title = "Trees vs Rangeland by Sub-Basin"
  )
```

## Sub-Basin Ranking by Tree Loss

Sub-basins ranked by the change in Trees and Rangeland cover between 2017 and 2023. Negative tree delta indicates loss; positive rangeland delta indicates gain.

```{r plot-lulc-ranking-ag, fig.cap="Sub-basins ranked by tree cover loss vs agriculture gain (2017-2023).", fig.width=10, fig.height=6}
ag_classes <- c("Crops", "Rangeland", "Bare Ground")

# Group agriculture and compute delta
ranking_ag <- lulc_all |>
  mutate(
    group = case_when(
      class_name == "Trees" ~ "Trees",
      class_name %in% ag_classes ~ "Agriculture",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(group), year %in% c(2017, 2023)) |>
  group_by(subbasin, year, group) |>
  summarize(pct = sum(pct), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = pct, names_prefix = "y") |>
  mutate(delta = y2023 - y2017) |>
  select(subbasin, group, delta) |>
  pivot_wider(names_from = group, values_from = delta) |>
  mutate(net_shift = Agriculture - Trees) |>
  arrange(desc(net_shift))

ranking_ag |>
  pivot_longer(cols = c(Trees, Agriculture), names_to = "class", values_to = "delta") |>
  mutate(subbasin = factor(subbasin, levels = ranking_ag$subbasin)) |>
  ggplot(aes(x = subbasin, y = delta, fill = class)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Trees" = "#006400", "Agriculture" = "#c0392b"), name = NULL) +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    y = "Change in % of Floodplain (2017-2023)",
    x = NULL,
    title = "Tree Loss vs Agriculture Gain (Crops + Rangeland + Bare Ground)"
  )
```

```{r plot-lulc-ranking-range, fig.cap="Sub-basins ranked by tree cover loss vs rangeland gain (2017-2023).", fig.width=10, fig.height=6}
ranking_range <- lulc_delta |>
  filter(class_name %in% c("Trees", "Rangeland")) |>
  select(subbasin, class_name, delta) |>
  pivot_wider(names_from = class_name, values_from = delta) |>
  mutate(net_shift = Rangeland - Trees) |>
  arrange(desc(net_shift))

ranking_range |>
  pivot_longer(cols = c(Trees, Rangeland), names_to = "class_name", values_to = "delta") |>
  mutate(subbasin = factor(subbasin, levels = ranking_range$subbasin)) |>
  ggplot(aes(x = subbasin, y = delta, fill = class_name)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Trees" = "#006400", "Rangeland" = "#FFBB22"), name = NULL) +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    y = "Change in % of Floodplain (2017-2023)",
    x = NULL,
    title = "Tree Loss vs Rangeland Gain Only"
  )
```

```{r plot-lulc-ranking-ag-pct, fig.cap="Sub-basins ranked by tree loss vs agriculture gain in percent of floodplain (2017-2023).", fig.width=10, fig.height=6}
ag_classes <- c("Crops", "Rangeland", "Bare Ground")

ranking_ag_pct <- lulc_all |>
  mutate(
    group = case_when(
      class_name == "Trees" ~ "Trees",
      class_name %in% ag_classes ~ "Agriculture",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(group), year %in% c(2017, 2023)) |>
  group_by(subbasin, year, group) |>
  summarize(pct = sum(pct), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = pct, names_prefix = "y") |>
  mutate(delta = y2023 - y2017) |>
  select(subbasin, group, delta) |>
  pivot_wider(names_from = group, values_from = delta) |>
  mutate(net_shift = Agriculture - Trees) |>
  arrange(desc(net_shift))

ranking_ag_pct |>
  pivot_longer(cols = c(Trees, Agriculture), names_to = "class", values_to = "delta") |>
  mutate(subbasin = factor(subbasin, levels = ranking_ag_pct$subbasin)) |>
  ggplot(aes(x = subbasin, y = delta, fill = class)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Trees" = "#006400", "Agriculture" = "#c0392b"), name = NULL) +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    y = "Change in % of Floodplain (2017-2023)",
    x = NULL,
    title = "Tree Loss vs Agriculture Gain (Crops + Rangeland + Bare Ground)"
  )
```

```{r plot-lulc-ranking-ag-ha, fig.cap="Sub-basins ranked by tree loss vs agriculture gain in hectares (2017-2023).", fig.width=10, fig.height=6}
ranking_ag_ha <- lulc_all |>
  mutate(
    group = case_when(
      class_name == "Trees" ~ "Trees",
      class_name %in% ag_classes ~ "Agriculture",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(group), year %in% c(2017, 2023)) |>
  group_by(subbasin, year, group) |>
  summarize(area = sum(area), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = area, names_prefix = "y") |>
  mutate(delta = y2023 - y2017) |>
  select(subbasin, group, delta) |>
  pivot_wider(names_from = group, values_from = delta) |>
  mutate(net_shift = Agriculture - Trees) |>
  arrange(desc(net_shift))

ranking_ag_ha |>
  pivot_longer(cols = c(Trees, Agriculture), names_to = "class", values_to = "delta") |>
  mutate(subbasin = factor(subbasin, levels = ranking_ag_ha$subbasin)) |>
  ggplot(aes(x = subbasin, y = delta, fill = class)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Trees" = "#006400", "Agriculture" = "#c0392b"), name = NULL) +
  geom_hline(yintercept = 0, linewidth = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    y = "Change in Area (ha, 2017-2023)",
    x = NULL,
    title = "Tree Loss vs Agriculture Gain — Absolute Area"
  )
```

## Agriculture Grouping

IO LULC classifies Crops, Rangeland, and Bare Ground as separate classes, but at 10 m resolution from single-date Sentinel-2 imagery these can represent different phases of the same agricultural land use — actively growing crops, fallow/grazed pasture, and freshly tilled soil. If the combined "Agriculture" proportion is stable while the individual classes swap, the apparent transitions are classification noise rather than real land cover change.

```{r plot-lulc-ag-group, fig.cap="Agriculture superclass (Crops + Rangeland + Bare Ground) vs Trees by sub-basin.", fig.width=12, fig.height=6}
ag_classes <- c("Crops", "Rangeland", "Bare Ground")

lulc_grouped <- lulc_all |>
  mutate(
    group = case_when(
      class_name == "Trees" ~ "Trees",
      class_name %in% ag_classes ~ "Agriculture",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(group)) |>
  group_by(subbasin, year, group) |>
  summarize(pct = sum(pct), .groups = "drop")

lulc_grouped |>
  mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = pct, color = group, group = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~subbasin, nrow = 2) +
  scale_color_manual(values = c("Trees" = "#006400", "Agriculture" = "#c0392b"), name = NULL) +
  theme_minimal() +
  labs(
    y = "% of Floodplain",
    x = NULL,
    title = "Trees vs Agriculture (Crops + Rangeland + Bare Ground)"
  )
```

```{r plot-lulc-ag-components, fig.cap="Individual agriculture components by sub-basin over time.", fig.width=12, fig.height=6}
lulc_all |>
  filter(class_name %in% ag_classes) |>
  mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = pct, color = class_name, group = class_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~subbasin, nrow = 2) +
  scale_color_manual(
    values = c("Crops" = "#E49635", "Rangeland" = "#FFBB22", "Bare Ground" = "#A59B8F"),
    name = NULL
  ) +
  theme_minimal() +
  labs(
    y = "% of Floodplain",
    x = NULL,
    title = "Agriculture Components by Sub-Basin"
  )
```

```{r lulc-ag-stability}
# Compare stability of grouped vs individual classes
ag_stability <- lulc_all |>
  mutate(
    group = case_when(
      class_name %in% ag_classes ~ "Agriculture (grouped)",
      TRUE ~ class_name
    )
  ) |>
  group_by(subbasin, year, group) |>
  summarize(pct = sum(pct), .groups = "drop") |>
  filter(group %in% c("Agriculture (grouped)", "Crops", "Rangeland", "Bare Ground", "Trees")) |>
  group_by(subbasin, group) |>
  summarize(
    pct_range = round(max(pct) - min(pct), 1),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = group, values_from = pct_range) |>
  arrange(subbasin)
```

```{r tab-lulc-ag-stability-cap, results="asis", eval=gitbook_on}
my_caption <- "Range of % floodplain cover across years (2017-2023) for individual classes vs grouped Agriculture. A small range for the grouped class with large ranges for components suggests classification swapping."

my_tab_caption()
```

```{r tab-lulc-ag-stability, eval=gitbook_on}
ag_stability |>
  my_dt_table(cols_freeze_left = 1, page_length = 10)
```

```{r tab-lulc-ag-stability-static, eval=identical(gitbook_on, FALSE)}
knitr::kable(ag_stability, digits = 1,
  caption = "Range of % floodplain cover across years (2017-2023) for individual classes vs grouped Agriculture.")
```

## Class Transitions

To understand whether tree loss is primarily converting to rangeland or also shifting to other classes (e.g., Built Area, Crops), we examine all class changes across the study period.

```{r plot-lulc-all-delta, fig.cap="Change in land cover by class and sub-basin (2017-2023).", fig.width=12, fig.height=8}
lulc_delta |>
  filter(abs(delta) >= 0.5) |>
  mutate(subbasin = factor(subbasin)) |>
  ggplot(aes(x = reorder(class_name, delta), y = delta, fill = delta > 0)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~subbasin, scales = "free_x") +
  scale_fill_manual(values = c("TRUE" = "#FFBB22", "FALSE" = "#006400")) +
  coord_flip() +
  theme_minimal() +
  labs(
    y = "Change in % of Floodplain (2017-2023)",
    x = NULL,
    title = "All Land Cover Class Changes by Sub-Basin"
  )
```

```{r lulc-transition-summary}
# Aggregate across all sub-basins to see overall class shifts
overall_delta <- lulc_delta |>
  group_by(class_name) |>
  summarize(
    mean_delta = round(mean(delta, na.rm = TRUE), 2),
    min_delta = round(min(delta, na.rm = TRUE), 2),
    max_delta = round(max(delta, na.rm = TRUE), 2),
    n_gaining = sum(delta > 0.5, na.rm = TRUE),
    n_losing = sum(delta < -0.5, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(mean_delta)
```

```{r tab-lulc-transitions-cap, results="asis", eval=gitbook_on}
my_caption <- "Mean land cover change (% of floodplain, 2017-2023) across all sub-basins. Negative values indicate loss."

my_tab_caption()
```

```{r tab-lulc-transitions, eval=gitbook_on}
overall_delta |>
  my_dt_table(cols_freeze_left = 1, page_length = 15)
```

```{r tab-lulc-transitions-static, eval=identical(gitbook_on, FALSE)}
knitr::kable(overall_delta, digits = 2,
  caption = "Mean land cover change (% of floodplain, 2017-2023) across all sub-basins. Negative values indicate loss.")
```

## Land Cover Composition

```{r tab-lulc-wide-cap, results="asis", eval=gitbook_on}
my_caption <- "Land cover composition (% of floodplain) by sub-basin and year."

my_tab_caption()
```

```{r tab-lulc-wide, eval=gitbook_on}
wide |>
  my_dt_table(cols_freeze_left = 2, page_length = 25)
```

```{r tab-lulc-wide-static, eval=identical(gitbook_on, FALSE)}
knitr::kable(wide, digits = 1,
  caption = "Land cover composition (% of floodplain) by sub-basin and year.")
```
