

# **Appendix 2 - Effectiveness Monitoring Data and Potential Restoration Site Review** {-#app-eff-mon}

<!--
To update this table:
1. Ensure QGIS project (restoration_wedzin_kwa) is synced and up to date locally
2. Run scripts/gis/amalgamate_field_forms.R to regenerate:
   - /Users/airvine/Projects/gis/restoration_wedzin_kwa/data_field/sites_reviewed_2024_202506.geojson
   - data/gis/sites_reviewed_2024_202506.geojson
3. Rebuild the report
-->

Site locations are available as a downloadable [`sites_reviewed_2024_202506.geojson`](https://github.com/NewGraphEnvironment/restoration_wedzin_kwa_2024/blob/main/data/gis/sites_reviewed_2024_202506.geojson) which renders as an interactive map on GitHub.

```{r tab-ree1-caption, results="asis"}
# Table \@ref(tab:tab-ree1-caption)
my_caption <- ("Results from effectiveness monitoring data collected in 2024 and 2025")
my_tab_caption()
```

```{r citation-convert, eval = FALSE}
# clean up the citations -  just doing this once bc it is really slow!!!
path_gis <- "/Users/airvine/Projects/gis/restoration_wedzin_kwa/data_field/sites_reviewed_2024_202506.geojson"
path_repo <- "data/gis/sites_reviewed_2024_202506.geojson"

dat <- sf::st_read(
  path_repo
)

dat |> 
  sf::st_drop_geometry() |> 
  dplyr::mutate(
    date = as.Date(date_time_start)
  ) |> 
  dplyr::select(
    # date_time_start,
    site_id,
    date,
    reference = citation_key,
    uav_flight,
    works_completed,
    new_site,
    assessment_comment
  ) |> 
  xciter::xct_keys_to_inline_table_col(col_format = "reference") |> 
  #   clean up when NA ends up in citation column
  dplyr::mutate(
    reference = dplyr::if_else(
      reference == "NA",
      "",
      reference
    )
  ) |> 
  # purrr::set_names(~ tools::toTitleCase(gsub("_", " ", .x)))
  readr::write_csv("data/inputs_extracted/monitoring_2024_2025.csv", na = "")

```


```{r tab-ree1}

dat_raw <- readr::read_csv(
  "data/inputs_extracted/monitoring_2024_2025.csv"
) 

label_short <- function(x) {
  x |>
    gsub("bulkley_meints_", "", x = _) |>
    gsub("bulkley_", "", x = _) |>
    gsub("-2021$", "", x = _) |>
    gsub("_0(\\d)", " \\1", x = _) |>
    gsub("_", " ", x = _) |>
    tools::toTitleCase() |>
    gsub("Mickilligan Road", "McKilligan", x = _) |>
    gsub("Mckilligan Rd", "McKilligan", x = _) |>
    gsub("Foxy Maxan Confluence", "Foxy-Maxan", x = _)
}

dat <- dat_raw |>
  dplyr::mutate(
    label = label_short(site_id),
    type = dplyr::case_when(
      works_completed == "yes" &
        !grepl("mud|kenneth", site_id, ignore.case = TRUE) ~
        "Effectiveness Monitoring",
      
      works_completed == "yes" &
        grepl("mud|kenneth", site_id, ignore.case = TRUE) ~
        "Fraser Site",
      
      
      TRUE ~ "Potential Site"
    ),
    type = factor(type, levels = c("Effectiveness Monitoring", "Potential Site", "Fraser Site"))
  ) |>
  dplyr::mutate(
    assessment_comment = dplyr::case_when(
      reference == "Gottesfeld and Rabnett (2007)" ~
        paste0("Newly proposed site adjacent to traditional fishing site. ", assessment_comment),
      new_site == "yes" ~
        paste0("Newly proposed site. ", assessment_comment),
      TRUE ~ assessment_comment
    ),
    uav_flight = dplyr::if_else(
      uav_flight == "No",
      NA_character_,
      uav_flight
    ) 
  )|> 
  dplyr::arrange(type, label) |>
  dplyr::select(
    label,
    date,
    type,
    reference,
    uav_flight,
    assessment_comment
  ) |> 
  tibble::rowid_to_column()

dat |>
  my_dt_table(cols_freeze_left = 2, page_length = 5)


```


```{r photo-setup, include=FALSE}
dir_photos <- "data/photos/sites"
path_sites_geojson <- "data/gis/sites_reviewed_2024_202506.geojson"

sites_all <- sf::st_read(path_sites_geojson, quiet = TRUE) |>
  dplyr::distinct(site_id, .keep_all = TRUE)
```


## Effectiveness Monitoring {-}

### BR1 — Bulkley River {-}

```{r site-br1-summary}
site <- dplyr::filter(sites_all, site_id == "BR1-2021")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-br1-photos, results='asis'}
photo_render("BR1-2021", dir_photos)
```

### BR2 — Bulkley River {-}

```{r site-br2-summary}
site <- dplyr::filter(sites_all, site_id == "BR2-2021")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-br2-photos, results='asis'}
photo_render("BR2-2021", dir_photos)
```

### BR5 — Bulkley River {-}

```{r site-br5-summary}
site <- dplyr::filter(sites_all, site_id == "BR5")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-br5-photos, results='asis'}
photo_render("BR5", dir_photos)
```

### Buc207 — Buck Creek {-}

```{r site-buc207-summary}
site <- dplyr::filter(sites_all, site_id == "Buc207")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-buc207-photos, results='asis'}
photo_render("Buc207", dir_photos)
```

### Craker Rd Lower — Bulkley River {-}

```{r site-craker-lower-summary}
site <- dplyr::filter(sites_all, site_id == "bulkley_meints_craker_rd_lower")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-craker-lower-photos, results='asis'}
photo_render("bulkley_meints_craker_rd_lower", dir_photos)
```

### McKilligan Upper — Bulkley River {-}

```{r site-mckilligan-upper-summary}
site <- dplyr::filter(sites_all, site_id == "Mickilligan Road Upper")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-mckilligan-upper-photos, results='asis'}
photo_render("Mickilligan Road Upper", dir_photos)
```

### MX1 — Maxan Creek {-}

```{r site-mx1-summary}
site <- dplyr::filter(sites_all, site_id == "MX1")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-mx1-photos, results='asis'}
photo_render("MX1", dir_photos)
```

### MX2 — Maxan Creek {-}

```{r site-mx2-summary}
site <- dplyr::filter(sites_all, site_id == "MX2")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-mx2-photos, results='asis'}
photo_render("MX2", dir_photos)
```


## Potential Site {-}

### Buc172 — Buck Creek {-}

```{r site-buc172-summary}
site <- dplyr::filter(sites_all, site_id == "Buc172")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-buc172-photos, results='asis'}
photo_render("Buc172", dir_photos)
```

### Buc7 — Buck Creek {-}

```{r site-buc7-summary}
site <- dplyr::filter(sites_all, site_id == "buc7")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-buc7-photos, results='asis'}
photo_render("buc7", dir_photos)
```

### Bul32 — Bulkley River {-}

```{r site-bul32-summary}
site <- dplyr::filter(sites_all, site_id == "bul32")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-bul32-photos, results='asis'}
photo_render("bul32", dir_photos)
```

### Bul38 — Bulkley River {-}

```{r site-bul38-summary}
site <- dplyr::filter(sites_all, site_id == "Bul38")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-bul38-photos, results='asis'}
photo_render("Bul38", dir_photos)
```

### Craker Rd Middle — Bulkley River {-}

```{r site-craker-middle-summary}
site <- dplyr::filter(sites_all, site_id == "bulkley_meints_craker_rd_middle")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-craker-middle-photos, results='asis'}
photo_render("bulkley_meints_craker_rd_middle", dir_photos)
```

### Craker Rd Upper — Bulkley River {-}

```{r site-craker-upper-summary}
site <- dplyr::filter(sites_all, site_id == "bulkley_meints_craker_rd_upper")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-craker-upper-photos, results='asis'}
photo_render("bulkley_meints_craker_rd_upper", dir_photos)
```

### Foxy-Maxan — Foxy Creek {-}

```{r site-foxy-maxan-summary}
site <- dplyr::filter(sites_all, site_id == "Foxy Maxan Confluence")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-foxy-maxan-photos, results='asis'}
photo_render("Foxy Maxan Confluence", dir_photos)
```

### McKilligan Middle — Bulkley River {-}

```{r site-mckilligan-middle-summary}
site <- dplyr::filter(sites_all, site_id == "Mckilligan Rd middle")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-mckilligan-middle-photos, results='asis'}
photo_render("Mckilligan Rd middle", dir_photos)
```

### Meints 1 — Bulkley River {-}

```{r site-meints1-summary}
site <- dplyr::filter(sites_all, site_id == "Meints_01")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-meints1-photos, results='asis'}
photo_render("Meints_01", dir_photos)
```

### Wilson 1 — Bulkley River {-}

```{r site-wilson1-summary}
site <- dplyr::filter(sites_all, site_id == "bulkley_wilson_01")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-wilson1-photos, results='asis'}
photo_render("bulkley_wilson_01", dir_photos)
```

### Wilson 2 — Bulkley River {-}

```{r site-wilson2-summary}
site <- dplyr::filter(sites_all, site_id == "bulkley_wilson_02")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-wilson2-photos, results='asis'}
photo_render("bulkley_wilson_02", dir_photos)
```

### Wilson 3 — Bulkley River {-}

```{r site-wilson3-summary}
site <- dplyr::filter(sites_all, site_id == "bulkley_wilson_03")
viewer_url <- stac_uav_viewer(site)
site_summary_table(site)
```

`r if (!is.na(viewer_url)) paste0("[UAV Orthophoto Viewer](", viewer_url, "){target='_blank'}")`

```{r site-wilson3-photos, results='asis'}
photo_render("bulkley_wilson_03", dir_photos)
```


