

# **Appendix 3 - Skeena Knowledge Trust - Data Available for Download ** {-}


A filtered subset of project packages (reports and associated files) available on SKT are listed in Table \@ref(tab:tab-skt-pkgs-cap). The table was created using [`/scripts/api_skt.R`](https://github.com/NewGraphEnvironment/restoration_wedzin_kwa_2024/tree/main/scripts/api_skt.R). Custom `R` function `fetch_package` can be used to download packages and is detailed in  [`/scripts/functions.R`](https://github.com/NewGraphEnvironment/restoration_wedzin_kwa_2024/tree/main/scripts/functions.R)

```{r tab-skt-pkgs-cap, results="asis"}
my_caption <- ("Skeena Knowledge Trust - Packages Available for Download")
my_tab_caption()
```


```{r tab-skt-pkgs}

path_skt_pkgs <- "data/skt/data_deets_20250217.rds"
# read iin the rds
skt_pkgs <- readRDS(path_skt_pkgs)

dataset_filter_raw <- "Old Growth Geospatial Data
Old Growth
Legal Old Growth
Air Photo
Photo BC*
name Data Files - Description - Zipped folder containing all the data files used for analysis
name - R Scripts - Description - Zipped folder containing the R-scripts used for data analysis
Riparian
Riparian and In-Stream Assessment of the Bulkley River System
Restoration
Floodplain mapping
Ecological Function
Habitat Report Cards: Upper Bulkley CU
Habitat Report Cards: Upper Skeena CU
Conservation Unit snapshots: Upper Skeena
Conservation Unit snapshots: Upper Bulkley
Annual Knowledge Plan
Mature and Old Forests
Tier 1 Fish and Fish Habitat Assessment
SSAF State of the Value Report
Integrated Watershed Restoration Plan
Upper Bulkley and Morice Water and Salmon Sustainability Views
Upper Bulkley Fish and Aquatic Review Summary of Data
Combined Upper Bulkley Air Photos
UB River Riparian Restoration Project Summary 2021
Upper Bulkley River riparian restoration 2022
Upper Bulkley River Watershed Water Temperature Monitoring Program 2016-21 Data Report
Hourly Water Temperature by Site
Wet'suwet'en
Water Quality"

# Split the text into a vector of strings
dataset_filter <- strsplit(dataset_filter_raw, split = "\n")[[1]]


# filter packages with stringr to find the ones that contain any of the strings in text_filter
# skt_pkgs_filtered <- skt_pkgs |> 
#   dplyr::filter(sapply(name, function(x) any(stringr::str_detect(x, dataset_filter))))

# group by why it was filtered
skt_pkgs_filtered <- purrr::map(
  dataset_filter, 
  ~skt_pkgs |>  dplyr::filter(stringr::str_detect(name, .x))
) |> 
  purrr::set_names(dataset_filter) |> 
  # remove empty list items
  purrr::discard(~nrow(.x) == 0) |> 
  dplyr::bind_rows(.id = "group")

skt_pkgs_filtered |> 
  dplyr::mutate(url = ngr::ngr_str_link_url(url_base = url, anchor_text = url)) |> 
  dplyr::select(group, package_name, description, everything()) |> 
  my_dt_table( cols_freeze_left = 1, page_length = 5, escape = FALSE)

```

<br>

Table \@ref(tab:tab-skt-geo-cap) represents spatial data available form the SKT geoserver and was created using [`/scripts/gis/skt/skt_geoserver_info.R`](https://github.com/NewGraphEnvironment/restoration_wedzin_kwa_2024/tree/main/scripts/gis/skt/skt_geoserver_info.R). Data
can be downloaded using [`skt_geoserver_dl.R`](https://github.com/NewGraphEnvironment/restoration_wedzin_kwa_2024/tree/main/scripts/gis/skt/skt_geoserver_dl.R) (will be converted to function following further testing).

```{r skt-geo-update, eval = params$update_gis}
source("scripts/gis/skt/skt_geoserver_info.R")
```

```{r tab-skt-geo-cap, results="asis"}
my_caption <- ("Skeena Knowledge Trust - GIS Data Available for Query and Download")
my_tab_caption()
```


```{r tab-skt-geo}

skt_geo <- readr::read_csv(
  "data/inputs_extracted/skt_geoserver_info.csv"
)

skt_geo |> 
  dplyr::select(-layer_name_raw) |> 
  my_dt_table( cols_freeze_left = 1, page_length = 5)

```


