

# **Appendix 5 – Example of Potential Restoration Sites Prioritized** {-#sites-ranked}

A summary of potential restoration sites prioritized based on ranking criteria detailed in Table \@ref(tab:tab-gis-params-cap)
is presented in Table \@ref(tab:tab-sites-ranked-cap). Please note that data is preliminary with ranking parameters and 
weights evolving based on ongoing research and consultation.

```{r tab-sites-ranked-cap, results="asis"}
my_caption <- ("Example of potential restoration sites prioritized.")
my_tab_caption()
```

```{r sites-copy, eval=params$update_gis}
# copy over the sites from the gis project into the repo for reproducability
path_in <- "/Users/airvine/Projects/gis/restoration_wedzin_kwa/sites_prioritized.geojson"
path_out <- "data/gis/sites_prioritized.geojson"


# convert to wsg84 so displays natively on github
sf::st_read(path_in, quiet = TRUE) |>
  sf::st_transform(4326) |>
  sf::st_write(path_out, quiet = TRUE, delete_dsn = TRUE)

```


```{r tab-sites-ranked}
path <- "data/gis/sites_prioritized.geojson"
  
sf::st_read(path, quiet = TRUE) |> 
  dplyr::filter(source != "ncfdc_1998_riparian") |> 
  sf::st_drop_geometry() |> 
  my_dt_table(cols_freeze_left = 2, page_length = 5, escape = TRUE)

```



