


# **Attachment 1 - Riparian Prescription Example** {-}

An example riparian prescription was implemented and monitored using iPhone LiDAR to capture 3D models of vegetation growth over time. The prescription incorporates 1m spacing of red-osier dogwood cuttings harvested while dormant and soaked for 48hrs, sod removal, soil amendment (bone meal), installation of biodegradable weed competition barrier (cardboard), 8 inches of non-acidic mulch (fir and maple) and automated drip irrigation running for 30 minutes every second day during growing season. Considerations for [riparian setback widths](https://github.com/NewGraphEnvironment/restoration_framework/issues/2) and detailed [planting prescription templates](https://github.com/NewGraphEnvironment/restoration_framework/issues/27) — including soil preparation, browse protection, monitoring plot design, and UAV mapping protocols — are linked above. Figure \@ref(fig:model-riparian-caption) shows the site after 2 growing seasons (October 15, 2024) and Figure \@ref(fig:model2-riparian-caption) shows the site after 3 growing seasons (October 10, 2025).

```{r model-riparian}
model_url <- '<iframe src="https://sketchfab.com/models/7d5b8b148c5447a9a0410fed205aff1f/embed" scrolling="no" title="20241015_dogwoodmulchirrigation" width="100%" height="600" frameBorder ="0" allowfullscreen mozallowfullscreen="true" webkitallowfullscreen="true" allow="autoplay; fullscreen;xr-spatial-tracking"></iframe>'

knitr::asis_output(model_url)
```

```{r model-riparian-caption, fig.cap= my_caption, out.width = "0.01%"}
my_photo = 'fig/pixel.png'

my_caption = paste0('3D model of revegetation after 2 growing seasons (October 15, 2024).')

knitr::include_graphics(my_photo, dpi = NA)

```

```{r model2-riparian}
model_url <- '<iframe src="https://sketchfab.com/models/4b9b0ffd5e894e2fa8ae1689391c27b6/embed" scrolling="no" title="20251010_dogwoodmulchirrigation" width="100%" height="600" frameBorder ="0" allowfullscreen mozallowfullscreen="true" webkitallowfullscreen="true" allow="autoplay; fullscreen;xr-spatial-tracking"></iframe>'

knitr::asis_output(model_url)
```

```{r model2-riparian-caption, fig.cap= my_caption, out.width = "0.01%"}
my_photo = 'fig/pixel.png'

my_caption = paste0('3D model of revegetation after 3 growing seasons (October 10, 2025).')

knitr::include_graphics(my_photo, dpi = NA)

```
