---
title: "Process UAV Imagery"
date: "Created: 2025-07-27 | Updated: `r format(Sys.Date(), '%Y-%m-%d')`"
output: 
  html_document:
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, dpi=60, out.width = "100%")

options(scipen=999)
options(knitr.kable.NA = '--') #'--'
options(knitr.kable.NAN = '--')
```

Here we process the imagery from Matt Sakals to get mosaics of the individual flights that are adjacent to each other.  We 
chose 15cm resolution due do the current resolution and so that it is reasonable to retrieve via https from QGIS. 

<br>

Here is the resolution for `bulkley-mckilligan-barren`.

```{r}
dir_1 <- "/Volumes/backup_2022/backups/new_graph/uav_imagery/skeena/bulkley/2024/bulkley-mckilligan-barren"
files_in <- fs::path(
  dir_1,
  c(
    "fullres-0-0.tif", 
    "fullres-0-1.tif", 
    "fullres-1-0.tif", 
    "fullres-1-1.tif"
  )
)

files_in1 |> 
  purrr::map(
    ngr::ngr_spk_res
  )

```


<br>

Here is the resolution for `bulkley-byman-richfield`.
```{r}
files_in2 <- fs::path(
  "/Volumes/backup_2022/backups/new_graph/uav_imagery/skeena/bulkley/2024/bulkley-byman-richfield",
  c(
    "Groot_Ortho_88mm_v3-0-0.tif",
    "Groot_Ortho_88mm_v3-1-0.tif",
    "Groot_Ortho_88mm_v3-2-0.tif"
  )
)

files_in2 |> 
  purrr::map(
    ngr::ngr_spk_res
  )

```


```{r}
library(ngr)

file_out <- "/Volumes/backup_2022/backups/new_graph/uav_imagery/skeena/bulkley/2024/bulkley-mckilligan-barren/ortho.tif"
args <- ngr::ngr_spk_gdalwarp(
  path_in = files_in1,
  path_out = file_out,
  t_srs = "EPSG:32609",
  target_resolution = c(0.15, 0.15)
)



file_out2 <- "/Volumes/backup_2022/backups/new_graph/uav_imagery/skeena/bulkley/2024/bulkley-byman-richfield/ortho.tif"


args2 <- ngr::ngr_spk_gdalwarp(
  path_in = files_in2,
  path_out = file_out2,
  t_srs = "EPSG:32609",
  target_resolution = c(0.15, 0.15)
)

args_list <- list(
  args,
  args2
)

args_list |> 
  purrr::walk(
  ~ processx::run(
    command = "gdalwarp",
    args = .x,
    echo = TRUE
  )
)

```


We have processed all UAV imagery for the Skeena [here](https://github.com/NewGraphEnvironment/stac_uav_bc/tree/main/scripts)
and loaded to s3 as a STAC compliant database.

